Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-13 paper/manuscript.tex
Accentmarks fixes.
remove version info for now.
WPI for addressing subsampling issues.

However, this is not the only metric which can or should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}. 
This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.

@@

There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate. Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$. The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy. Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
\begin{eqnarray}
g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
\end{eqnarray}
where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
\begin{eqnarray}
\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
\end{eqnarray}
with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
\begin{eqnarray}
C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
\end{eqnarray}
The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).

@@

If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.

\paragraph{Cautions and considerations}
Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.

@@

The computation of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used. 
The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral. Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be a helpful diagnostic. 
Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently. 
Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.

>>>>

However, this is not the only metric which can or should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance requires sampling the different kinetically stable states~\cite{mobley2012perspective}. 
This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.

@@

There is one important thing to note about the uncertainty estimates when summing multiple free energy differnces together to calculate an overall estimate of a free energy difference. Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$. The variances of each of the free energies cannot simply be summed together into the variance of the overall free energy. Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the statistical uncertainty, as most estimates of the uncertainty are derived assuming unbiased samples.
To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
\begin{eqnarray}
g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
\end{eqnarray}
where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
\begin{eqnarray}
\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
\end{eqnarray}
with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
\begin{eqnarray}
C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
\end{eqnarray}
The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).

@@

However, samples uncorrelated by this procedure are only approximately
uncorrelated.  Recent careful investigation has shown that the lowest
uncertainty results can be obtained by using \textit{all} of the data
to obtain an estimate of the free energy itself, but using only
subsampled data to calculate error estimates using the formulas or by
using bootstrapping.  In fact, one can show that the standard
deviation of the estimate determined by data that has been subsampled
is the unbiased estimator of the free energy estimate using all the
data, but is actually an underestimate of the uncertainty of estimate
using just the subsampled data.

We don't need to use all of the data to recover the remaining
correlation.  In practice, if one subsamples the data with a
correlation interval of $g/10$, where $g$ is the correlation
time. Thus, if $g=3.0$ in units of timesteps, it is best to use all of
the data.  If $g=100.0$, then one loses essentially no information by
subsampling with $g=10.0$, and still saves significant time by not
using all of the now-completely duplicative data.

If independent simulations at each $\lambda$ are used, the alchemical
state $\vec{\lambda}$ may have a significant impact on the correlation
time, and these simulations should be subsampled independently using a
separate estimate of the statistical inefficiency $g$ for each
alchemical state. If coupled simulations are used (such as a
Hamiltonian replica exchange simulation), the replicas should undergo
equivalent random walks in alchemical space, and the replicas can be
subsampled with the same $g$ to generate an equal number of
uncorrelated samples at each alchemical state. Conveniently, the
approach described above for automated equilibration detection
produces an appropriate estimate of $g$ over the production region for
automating this process.

Our overall recommendations are therefore to determine the subsampling
period $g$, subsample data at $g/10$ (i.e. more frequently than
specified by $g$) to compute the free energy estimates, and then use
fully subsampled data (with subsampling interval $g$) to estimate the
uncertainties of the estimates.

\paragraph{Cautions and considerations}
Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge  \sim 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.

@@

Computating of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended. The trapezoid rule easily allows unequal spacing of $\vec{\lambda}$ states, which is often required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used if care is taken.  
Determining of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral. Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values, which can be done automatically with tool such as alchemlyb,  can be a helpful diagnostic. 
Additionally, computating the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently. 
Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\eref{subsec:estimators} for more discussion of $w_k$).
